Fast and exact visibility on digitized shapes and application to saliency-aware normal estimation

### 2025-04-25 DGMM2025/introduction.tex
typos fixing

    Visibility is a fundamental concept in computational geometry and
    digital topology, with applications ranging from computer vision
    (occlusions), geometric modeling (feature detection,
    geodesics) to computer graphics (path tracing,
    shadowmap). Visibility within polygons in the plane has been
    extensively studied in the litterature~\cite{ghosh:2007-book} and
    is still studied in higher dimensional
    spaces~\cite{orourke:2017-book}. The general problem has a high
    complexity. For instance, the 3d visibility complex between $n$
    spheres and occluding spheres has a complexity
    $O(n^4)$~\cite{durand:2002-tog}. Exact 3d visibility between
    polygons and occluding polygons is often casted into 5d Plücker
    space~\cite{nirenstein:2002-ewr}, which nicely represents 3d lines
    and their respective positions. The computational cost remains
    high even with decomposition into coarse cells for speed up
    (several days of computation for 1M triangles at the time).

    To increase efficiency, several authors have casted the
    visibility problem into a digital space, generally $\Z^3$ or its
    decomposition into cubical cells. Visibility is simplified as the
    question whether there is a connected digital straight line
    joining a pair of cells without occluding cell(s)
    (e.g. Soille~\cite{soille:1994-prl} uses Bresenham
    segments). Coeurjolly \emph{et al.}~\cite{coeurjolly:2004-prl}
    uses a non-symmetric visibility definition and preimage
    computations to speed up these
    algorithms. Chica~\cite{chica:2008-spm} uses a half-cell erosion
    of the complementary space and Bresenham lines to decide
    visibility.

    We propose an algorithm that solves the following visibility
    problem: given a digital set $X \subset \Z^d$, two points $p,q$ of
    $X$ are \emph{visible} whenever the Euclidean straight line
    segment $\lbrack p,q \rbrack$ is never at chessboard distance
    ($\infty$-distance) greater or equal to $1$ from $X$. It is
    equivalent to the \emph{cotangency} of $p$ and $q$ in $X$, as
    formalized in~\cite{lachaud:2021-dgmm,lachaud:2022-jmiv}, and can
    be viewed as an inclusion problem between sets of lattice points
    in this form. Indeed cells surounding $X$ can be encoded as
    lattice points (Khalimsky coding), as well as the cells covering
    lattice directions. The originality of our method is to encode
    subsets of lattice points as sets of integer intervals and to
    reduce all the inclusion problems as intersections and
    translations of integer intervals. For a given lattice vector
    $\mathbf{t}$, we solve the visibility of every pair of points
    $(p,p+\mathbf{t})$ of $X$. Global visibility is achieved by
    testing all sought directions within a given range, and is
    parallelized straightforwardly.

>>>>

    Visibility is a fundamental concept in computational geometry and
    digital topology, with applications ranging from computer vision
    (occlusions), geometric modeling (feature detection,
    geodesics) to computer graphics (path tracing,
    shadowmap). Visibility within polygons in the plane has been
    extensively studied in the litterature~\cite{ghosh:2007-book} and
    is still studied in higher dimensional
    spaces~\cite{orourke:2017-book}. The general problem has a high
    complexity. For instance, the 3d visibility complex between $n$
    spheres and occluding spheres has a complexity
    $O(n^4)$~\cite{durand:2002-tog}. Exact 3d visibility between
    polygons and occluding polygons is often cast into 5d Plücker
    space~\cite{nirenstein:2002-ewr}, which nicely represents 3d lines
    and their respective positions. The computational cost remains
    high even with decomposition into coarse cells for speed up
    (several days of computation for 1M triangles at the time).

    To increase efficiency, several authors have cast the
    visibility problem into a digital space, generally $\Z^3$ or its
    decomposition into cubical cells. Visibility is simplified as
    whether there is a connected digital straight line
    joining a pair of cells without occluding cell(s)
    (e.g.\ Soille~\cite{soille:1994-prl} uses Bresenham
    segments). Coeurjolly \emph{et al.}~\cite{coeurjolly:2004-prl}
    uses a non-symmetric visibility definition and preimage
    computations to speed up these
    algorithms. Chica~\cite{chica:2008-spm} uses a half-cell erosion
    of the complementary space and Bresenham lines to decide
    visibility.

    We propose an algorithm that solves the following visibility
    problem: given a digital set $X \subset \Z^d$, two points $p,q$ of
    $X$ are \emph{visible} whenever the Euclidean straight line
    segment $\lbrack p,q \rbrack$ is never at chessboard distance
    ($\infty$-distance) greater or equal to $1$ from $X$. It is
    equivalent to the \emph{cotangency} of $p$ and $q$ in $X$, as
    formalized in~\cite{lachaud:2021-dgmm,lachaud:2022-jmiv}, and can
    be viewed as an inclusion problem between sets of lattice points
    in this form. Indeed, cells surounding $X$ can be encoded as
    lattice points (Khalimsky coding), as well as the cells covering
    lattice directions. The originality of our method is to encode
    subsets of lattice points as sets of integer intervals and to
    reduce all the inclusion problems as intersections and
    translations of integer intervals. For a given lattice vector
    $\mathbf{t}$, we solve the visibility of every pair of points
    $(p,p+\mathbf{t})$ of $X$. Global visibility is achieved by
    testing all sought directions within a given range, and is
    parallelized straightforwardly.

### 2025-04-25 DGMM2025/normalsandcurvature.tex
typos fixing

\paragraph{Experimental validation.}
Figure~\ref{fig:normals-estimation} shows an example
of our normals compared to normals computed with the integral invariant (II) estimator~\cite{Lachaud:2017-lnm}.
In order to compare quantitatively the different normal estimators, we compute the
root mean square error (RMSE) and the maximum error $E_{\max}$ of the
normals computed on a digital surface with respect to the normals computed
on the continuous surface. We use 4 different digital surfaces (ellipsoid,
goursat, rcube and sphere9). We see that the RMSE and $E_{\max}$ errors of our
estimator are convergent with respect to the grid resolution with a
rate of convergence of $h^{\frac{2}{3}}$ for the RMSE and $h^{\frac{1}{2}}$ for
$E_{\max}$.

@@

The CTriv estimator only performs averaging of trivial normal
directions along the surface. It behaves much better than II near
features (although it tends to smooth curvatures) since it does
not use information from across the gap. However some oscillations
of the curvatures are distinguishable especially along the flat
parts, and some curvature estimates are erroneous on smoother
parts (like the edge above 6 for $\kappa_2$ or several dice
vertices).

>>>>

\paragraph{Experimental validation.}
Figure~\ref{fig:normals-estimation} shows an example
of our normals compared to normals computed with the integral invariant (II) estimator~\cite{Lachaud:2017-lnm}.
In order to compare quantitatively the different normal estimators, we compute the
root-mean-square error (RMSE) and the maximum error $E_{\max}$ of the
normals computed on a digital surface with respect to the normals computed
on the continuous surface (Figure~\ref{fig:errors-normals}). We use 4 different digital surfaces (ellipsoid,
goursat, rcube and sphere9). We see that the RMSE and $E_{\max}$ errors of our
estimator are convergent with respect to the grid resolution with a
rate of convergence of $h^{\frac{2}{3}}$ for the RMSE and $h^{\frac{1}{2}}$ for
$E_{\max}$.

@@

The CTriv estimator only performs averaging of trivial normal
directions along the surface. It behaves much better than II near
features (although it tends to smooth curvatures) since it does
not use information from across the gap. However, some oscillations
of the curvatures are distinguishable especially along the flat
parts, and some curvature estimates are erroneous on smoother
parts (like the edge above 6 for $\kappa_2$ or several dice
vertices).

### 2025-04-25 DGMM2025/visibility.tex
typos fixing

    The \emph{topological closure} of a cell $\tau$ is denoted by
    $\bar{\tau}$. The \emph{star} of a cell $\sigma$ is the subcomplex
    $\Star{\sigma}:=\{\tau \in \mathcal{C}, \sigma \subset \bar{\tau}
    \}$. It is naturally extended to any subcomplex $K$ by union,
    i.e. $\Star{K} := \cup_{\sigma \in K} \Star{\sigma}$, and more
    generally to any subset $Y$ of $\R^d$ as $\Star{Y}:=\{ c \in \C,
    \bar{c} \cap Y \neq \emptyset \}$. The \emph{body} $\|K\|$ of $K$
    is its realization in $\R^d$, i.e. $\|K\|:=\cup_{c \in K} c$. One
    can check that $\Star{K}=\Star{\|K\|}$, so these definitions are
    sound.

    %% A cell $\sigma$ is a face of another cell $\tau$ whenever $\sigma$
    %% is a subset of the topological closure $\bar{\tau}$ of $\tau$,
    %% written $\sigma \preccurlyeq \tau$.  Given any subcomplex K of
    %% $\mathcal{C}^d$, the star $Star(K)$ of K is $\{\tau \in
    %% \mathcal{C}^d, s.t.\ \exists\sigma \in K,\sigma \preccurlyeq
    %% \tau\}$.
    \begin{definition}[Visibility]
      Let $Z \subset \Z^d$ be a digital set. Two digital points $p$
      and $q$ are \emph{visible in $Z$} if and only if the straight
      segment $[p, q]$ is included in $\Star{Z}$, i.e., $\Star{[p, q]}
      \subseteq \Star{Z}$.
    \end{definition}
    This definition corresponds to the concept of tangency presented
    in \cite{lachaud:2022-jmiv}, limited to pair of points. One can
    check that $p$ and $q$ must belong to $Z$ to be visible.
    Figure~\ref{fig:visibility-2d} illustrates what is considered
    visible or not on some examples.

@@

    This characteristic has implications for the applicability of
    \cite[Algorithm 3]{lachaud:2022-jmiv}, which is a breadth-first
    like algorithm for computing all the points of $X$ visible in $X$
    to a source point $p$. This algorithm sometimes outputs an
    incomplete collection of visible points. This is why in the next
    section we present an algorithm that computes the exact visibility
    pairs within $X$, without recourse to connectedness in its formulation.

>>>>

    The \emph{topological closure} of a cell $\tau$ is denoted by
    $\bar{\tau}$. The \emph{star} of a cell $\sigma$ is the subcomplex
    $\Star{\sigma}\coloneqq\{\tau \in \mathcal{C}, \sigma \subset \bar{\tau}
    \}$. It is naturally extended to any subcomplex $K$ by union,
    i.e. $\Star{K} \coloneqq \cup_{\sigma \in K} \Star{\sigma}$, and more
    generally to any subset $Y$ of $\R^d$ as $\Star{Y}\coloneqq\{ c \in \C,
    \bar{c} \cap Y \neq \emptyset \}$. The \emph{body} $\|K\|$ of $K$
    is its realization in $\R^d$, i.e. $\|K\|\coloneqq\cup_{c \in K} c$. One
    can check that $\Star{K}=\Star{\|K\|}$, so these definitions are
    sound.

    %% A cell $\sigma$ is a face of another cell $\tau$ whenever $\sigma$
    %% is a subset of the topological closure $\bar{\tau}$ of $\tau$,
    %% written $\sigma \preccurlyeq \tau$.  Given any subcomplex K of
    %% $\mathcal{C}^d$, the star $Star(K)$ of K is $\{\tau \in
    %% \mathcal{C}^d, s.t.\ \exists\sigma \in K,\sigma \preccurlyeq
    %% \tau\}$.
    \begin{definition}[Visibility]
      Let $Z \subset \Z^d$ be a digital set. Two digital points $p$
      and $q$ are \emph{visible in $Z$} if and only if the straight
      segment $[p, q]$ is included in $\Star{Z}$, i.e., $\Star{[p, q]}
      \subseteq \Star{Z}$.
    \end{definition}
    This definition corresponds to the concept of tangency presented
    in~\cite{lachaud:2022-jmiv}, limited to pairs of points. One can
    check that $p$ and $q$ must belong to $Z$ to be visible.
    Figure~\ref{fig:visibility-2d} illustrates what is considered
    visible or not on some examples.

@@

    This characteristic has implications for the applicability of~\cite[Algorithm 3]{lachaud:2022-jmiv}, which is a breadth-first
    like algorithm for computing all the points of $X$ visible in $X$
    to a source point $p$. This algorithm sometimes outputs an
    incomplete collection of visible points. This is why in the next
    section we present an algorithm that computes the exact visibility
    pairs within $X$, without recourse to connectedness in its formulation.
